Scagnostics and Projection Pursuit

### 2018-01-30 Draft/Draft.tex
First commit

---
author:
  - name: Ursula Laa
    affiliation: Monash University
    address: >
      First line
      Second line
    email: \email{ursula.laa@monash.edu}
    url: http://rstudio.com
  - name: Second Author
    affiliation: Affiliation
title:
  formatted: "Scagnostics and Projection Pursuit"
  # If you use tex in the formatted title, also supply version without
  #plain:     "A Capitalized Title: Something about a Package foo"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [scagnostics, tour, projection pursuit]
  #plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, 
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = TRUE, external = TRUE, dev = "pdf",
  fig.height = 5, fig.width = 8, out.width = "\\textwidth"
)
read_chunk('src/main.R')
```

```{r load}
```

# Introduction

@@

Here we aim to extend the definition of interesting projections and available projection persuit indices by studying potential
new index functions.
As a starting point we consdier the scagnostics measures that were proposed for selecting bivariate scatterplots in high-dimensional datasets ADDREF (??).
Rather than considering the scagnostics measures for all possible bivariate combinations, we want to use one of the measures as an index function for a guided tour of 2 dimensional projections.
We first study the feasability of this approach and discuss potential pitfalls.
We then aim to define more robust measures ... ???

# Scagnostics measures

Eight scagnostics measures have been defined, the definitions are based on the convex and alpha hull as well as the minimum spanning tree.
Give definitions here...

@@

# Analysis setup

@@

# Application to particle physics data

## Data description

For the following illustrations we work with the particle physics dataset (ATLAS Collaboration, 2015).
This data includes a number of predictions for observables (in agreement with current limits) and the so-called finetuning measure for a large set of parameter points in a supersymmetric model.
The observables describe predictions for decay branching ratios and other precision observables that are generally used to constrain new models of particle physics.
We use the log transformed values for the variables XYZ but not the finetuning, and we rescale all variables to a [0,1] interval
to obtain comparable scales.
Contributions to the variable gminus2 can be negative or positive, we therefore add a small shift before the log transformation of this variable.

## Software setup

@@

## Scagnositics of grand tour projections

```{r scagnostics-summary, fig.height= 14, fig.width=9, fig.cap = "(ref:scagnostics-summary)"}
```

(ref:scagnostics-summary) Summary of evolution of scagnostics measures over 100 interpolated grand tour projections
We see that three of the considered measures, Clumpy, Sparse and Striated, are not relevant for the considered data and projections,
as the valuse are consistently low (clearly below 0.1).
The remaining measures show large variation depending on the projection, sample size and also often strong dependence on the
outlier removal before calculation of scagnostics measures.
Moreover, we note that despite considering interpolated projections, i.e. a smooth transition between the randomly selected projection planes in the grand tour, the variation of the scagnostics measures does not follow a smooth evolution but exhibits several spikes.
This behaviour is expected to result in difficulites when using scagnostics measures as a projection pursuit index in the context of a dynamic guided tour.
To further understand this behaviour let us study the projected data points and also the binned version for some of the projections exhibiting a spike.
We first show in Figure ADDREF projections t=18/19/20 of the small data set (500 data points).
Not that t=19 has been marked by a vertical line in Figure ADDREF.
The left column shows the projected data points, and the right column the binned data used to calculate the scagnositcs measures.
The scagnostics measures in the three projections are summarised in Table ADDREF.
We see that despite similar distribution of the points in all three projections, the outlying measure is similar for t=18 and t=20 but jumpig down for t=19.
Differences can however be observed in the binned data (right column of Figure ADDREF), likely causing the behaviour of the measures.

@@

```{r scatter-binning, fig.height=2, fig.cap = "(ref:scatter-binning)"}
```

```{r scatter-table}
```

@@

```{r scagnositcs-timing, fig.cap = "(ref:scagnostics-timing)"}
```

## Smooth transitions

We next turn to the question of how we can obtain smooth transitions of the scagnostics measures.
(Below just an example of using ggplot smoothing to display what kind of behaviour we want to find. It follows the general trend without capturing the spikes.)
```{r smoothing, fig.cap = "(ref:smooting)"}
```

@@

### Binning in high dimensions?
### Outlier removal before binning?

## Test with guided tour
try out scagnostics measures as guided tour index
FIXME still need to write script for this

## Additional indices to consider
FIXME load Katrins package and try the index functions she implemented

>>>>

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Ursula Laa\\Monash University \And Second Author\\Affiliation
}
\title{Scagnostics and Projection Pursuit}

\Plainauthor{Ursula Laa, Second Author}

\Abstract{
The abstract of the article.
}

\Keywords{scagnostics, tour, projection pursuit}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Ursula Laa\\
  Monash University\\
  First line Second line\\
  E-mail: \email{ursula.laa@monash.edu}\\
  URL: \url{http://rstudio.com}\\~\\
    }

\usepackage{amsmath}

\begin{document}

\section{Introduction}\label{introduction}

@@

Here we aim to extend the definition of interesting projections and
available projection persuit indices by studying potential new index
functions. As a starting point we consdier the scagnostics measures that
were proposed for selecting bivariate scatterplots in high-dimensional
datasets ADDREF (??). Rather than considering the scagnostics measures
for all possible bivariate combinations, we want to use one of the
measures as an index function for a guided tour of 2 dimensional
projections. We first study the feasability of this approach and discuss
potential pitfalls. We then aim to define more robust measures \ldots{}
???

\section{Scagnostics measures}\label{scagnostics-measures}

Eight scagnostics measures have been defined, the definitions are based
on the convex and alpha hull as well as the minimum spanning tree. Give
definitions here\ldots{}

@@

\section{Analysis setup}\label{analysis-setup}

@@

\section{Application to particle physics
data}\label{application-to-particle-physics-data}

\subsection{Data description}\label{data-description}

For the following illustrations we work with the particle physics
dataset (ATLAS Collaboration, 2015). This data includes a number of
predictions for observables (in agreement with current limits) and the
so-called finetuning measure for a large set of parameter points in a
supersymmetric model. The observables describe predictions for decay
branching ratios and other precision observables that are generally used
to constrain new models of particle physics. We use the log transformed
values for the variables XYZ but not the finetuning, and we rescale all
variables to a {[}0,1{]} interval to obtain comparable scales.
Contributions to the variable gminus2 can be negative or positive, we
therefore add a small shift before the log transformation of this
variable.

\subsection{Software setup}\label{software-setup}

@@

\subsection{Scagnositics of grand tour
projections}\label{scagnositics-of-grand-tour-projections}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{figure/scagnostics-summary-1}

}

\caption[(ref:scagnostics-summary)]{(ref:scagnostics-summary)}\label{fig:scagnostics-summary}
\end{figure}
\end{CodeChunk}

(ref:scagnostics-summary) Summary of evolution of scagnostics measures
over 100 interpolated grand tour projections We see that three of the
considered measures, Clumpy, Sparse and Striated, are not relevant for
the considered data and projections, as the valuse are consistently low
(clearly below 0.1). The remaining measures show large variation
depending on the projection, sample size and also often strong
dependence on the outlier removal before calculation of scagnostics
measures. Moreover, we note that despite considering interpolated
projections, i.e.~a smooth transition between the randomly selected
projection planes in the grand tour, the variation of the scagnostics
measures does not follow a smooth evolution but exhibits several spikes.
This behaviour is expected to result in difficulites when using
scagnostics measures as a projection pursuit index in the context of a
dynamic guided tour. To further understand this behaviour let us study
the projected data points and also the binned version for some of the
projections exhibiting a spike. We first show in Figure ADDREF
projections t=18/19/20 of the small data set (500 data points). Not that
t=19 has been marked by a vertical line in Figure ADDREF. The left
column shows the projected data points, and the right column the binned
data used to calculate the scagnositcs measures. The scagnostics
measures in the three projections are summarised in Table ADDREF. We see
that despite similar distribution of the points in all three
projections, the outlying measure is similar for t=18 and t=20 but
jumpig down for t=19. Differences can however be observed in the binned
data (right column of Figure ADDREF), likely causing the behaviour of
the measures.

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{figure/scatter-binning-1} \includegraphics[width=\textwidth]{figure/scatter-binning-2} \includegraphics[width=\textwidth]{figure/scatter-binning-3}

}

\caption[(ref:scatter-binning)]{(ref:scatter-binning)}\label{fig:scatter-binning}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}

\begin{CodeOutput}
                   s1          s2          s3
Outlying  0.275676305 0.163548110 0.270907044
Skewed    0.549038842 0.575422054 0.583569727
Clumpy    0.004123711 0.008163265 0.004175365
Sparse    0.038765964 0.042779580 0.040551326
Striated  0.059405941 0.060000000 0.037433155
Convex    0.611714549 0.602591732 0.546399191
Skinny    0.483884014 0.495557112 0.564627720
Stringy   0.424567811 0.368764181 0.329434667
Monotonic 0.077330543 0.094217913 0.103036354
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{figure/scagnositcs-timing-1}

}

\caption[(ref:scagnostics-timing)]{(ref:scagnostics-timing)}\label{fig:scagnositcs-timing}
\end{figure}
\end{CodeChunk}

\subsection{Smooth transitions}\label{smooth-transitions}

We next turn to the question of how we can obtain smooth transitions of
the scagnostics measures. (Below just an example of using ggplot
smoothing to display what kind of behaviour we want to find. It follows
the general trend without capturing the spikes.)

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{figure/smoothing-1}

}

\caption[(ref:smooting)]{(ref:smooting)}\label{fig:smoothing}
\end{figure}
\end{CodeChunk}

@@

\subsubsection{Binning in high
dimensions?}\label{binning-in-high-dimensions}

\subsubsection{Outlier removal before
binning?}\label{outlier-removal-before-binning}

\subsection{Test with guided tour}\label{test-with-guided-tour}

try out scagnostics measures as guided tour index FIXME still need to
write script for this

\subsection{Additional indices to
consider}\label{additional-indices-to-consider}

FIXME load Katrins package and try the index functions she implemented

\end{document}
